Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-04-10 Paper/manuscript_v2/manuscript_07_01.tex
edits from 5.3 to end

\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\DeclareMathOperator{\LogNormal}{LogNormal}

@@

We assume a GP prior for $f \sim \mathcal{GP}(\mu(\bm{x}), k(\bm{x}, \bm{x}'))$, where $\mu: {\rm I\!R}^D \rightarrow {\rm I\!R}$ and $k: {\rm I\!R}^D \times {\rm I\!R}^D \rightarrow {\rm I\!R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\bm{x}) &= \mathbb{E}\!\left[f(\bm{x})\right],\\ 
 	k(\bm{x}, \bm{x}') &= \mathbb{E}\!\left[\left( f(\bm{x}) - \mu(\bm{x}) \right)\left( f(\bm{x}') - \mu(\bm{x}') \right)\right].
\end{align*}

@@

\begin{figure}
\centering
%\subfigure{\includegraphics[scale=0.38, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig1_legend_01.pdf}}\\
%\vspace{-9mm}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 3mm 0mm, clip]{ch5_fig1_Cov_J.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig1_legend.pdf}}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the regular GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig2_Sigma_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 3mm 0mm, clip]{ch5_fig2_Cov_L.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig2_legend.pdf}}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the regular GP model (dashed red line) and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure}

Next, we analyze how the interaction effects between $m$ and $c$ affects the quality of the approximation. The lengthscale and marginal variance of the covariance function will no longer be fixed but instead their posterior marginal distributions estimated using dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual}, for both regular GP and HSGP models. Figure \ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the data for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the regular GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated lengthscale and marginal variance for the regular GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computations would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$  could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure \ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

@@

\begin{figure}
\centering
\begin{tabular}{ c c c }
\arrayrulecolor{darkgray}\hline
c = 1.05 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part1.pdf}
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part1.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part1.pdf} & 
\multirow{32}{1cm}{ \includegraphics[scale=0.35, trim = 28mm 30mm 100mm 30mm, clip]{ch5_fig3_legend.pdf}}\\ 
\arrayrulecolor{lightgray}\cline{1-2}
c = 1.1 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part2.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part2.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part2.pdf} &\\
\cline{1-2}
c = 1.2 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part3.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part3.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part3.pdf} &\\
\cline{1-2}
c = 1.5 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part4.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part4.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part4.pdf} & \\
\cline{1-2}
c = 2 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part5.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part5.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part5.pdf} & \\
\cline{1-2}
c = 2.5 &
\includegraphics[scale=0.215, trim = 0mm 4mm 0mm 14mm, clip]{ch5_fig3_Post_part6.pdf}
\includegraphics[scale=0.215, trim = 0mm 4mm 0mm 14mm, clip]{ch5_fig3_Sigma_part6.pdf}
\includegraphics[scale=0.215, trim = 0mm 4mm 5mm 14mm, clip]{ch5_fig3_Cov_part6.pdf} &\\
\arrayrulecolor{darkgray}\hline
\end{tabular}
\caption{Posterior mean predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the regular GP model and the HSGP model for different number of basis functions $m$ and for different values of the boundary factor $c$. The dashed red line of the regular GP can hardly be seen in the some of the plots because it is under the blue lines.}
  \label{fig3_Post_part1}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the regular GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig5_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 28mm 119mm 0mm, clip]{ch5_fig5_legend.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both regular GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure}

@@

In this example, we make a comparison of the lengthscale estimates
obtained from the regular GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.
For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both regular GP and HSGP models fitted to those datasets. The lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As noted previously, an accurate estimate of the lengthscale can be a good indicator of a close approximation of the HSGP model to the regular GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the regular GP models, as a function of the lengthscale and number of basis functions.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Mean posterior predictions of both regular GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). The dashed red line of the regular GP can hardly be seen in the plots because it is almost perfectly overlayed by the blue lines.}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ccc}
\multicolumn{3}{c}{ \includegraphics[scale=0.42, trim = 25mm 117mm 90mm 45mm, clip]{ch5_fig8_legend.pdf}}\\
\includegraphics[scale=0.31, trim = 0mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part1.pdf} & \hspace{-4mm}
\includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part2.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part3.pdf} \\ 
\includegraphics[scale=0.31, trim = 0mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part4.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part5.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part6.pdf}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional lengthscales ($\ell$), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding lengthscale estimates $(\hat{\ell})$ from the regular GP and HSGP models. 95\% confident intervals of the lengthscale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

@@

The regular GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(x) &\sim \mathcal{GP}(0, k(x, x', \theta)),
\end{split}
\end{align*}
where $I$ represents the identity matrix and $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function values to the noisy data. The previous formulation corresponds to the latent form of a GP model. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with a Mat{\'e}rn($\nu$=3/2) covariance function $k$. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$.

@@

In the HSGP model, the latent function values $f(x)$ are approximated as in equation (\ref{eq_approxf}), with the Mat{\'e}rn($\nu$=3/2) spectral density $S$ as in equation (\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively. 
In order to do model comparison, in addition to the regular GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the regular GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$; see equation (\ref{eq_boundary})), and the spline model with 80 knots. The true data-generative function and the noisy observations are also plotted. The sample observations are plotted as circles and the out-of-sample or test data, which have not been taking part on training the models, are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation. The posteriors of the three models, regular GP, HSGP and spline, are pretty similar within the interpolation input space. However, when extrapolating the spline model solution clearly differs from the regular GP and HSGP models as well as the actual observations.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{ch5_fig10_Posteriors_exI.pdf}
\caption{Posterior predictive means of the proposed HSGP model, the regular GP model, and the spline model. 95\% credible intervals are plotted as dashed lines.}
  \label{ch5_fig10_Posteriors_exI}
\end{figure}

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{ch5_fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{ch5_fig10_Posteriors_exI} and \ref{ch5_fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the regular GP model for interpolating and extrapolating data. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

>>>>

\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\DeclareMathOperator{\LogNormal}{LogNormal}
\DeclareMathOperator{\GP}{\mathcal{GP}}

@@

We assume a GP prior for $f \sim \GP(\mu(\bm{x}), k(\bm{x}, \bm{x}'))$, where $\mu: {\rm I\!R}^D \rightarrow {\rm I\!R}$ and $k: {\rm I\!R}^D \times {\rm I\!R}^D \rightarrow {\rm I\!R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\bm{x}) &= \mathbb{E}\!\left[f(\bm{x})\right],\\ 
 	k(\bm{x}, \bm{x}') &= \mathbb{E}\!\left[\left( f(\bm{x}) - \mu(\bm{x}) \right)\left( f(\bm{x}') - \mu(\bm{x}') \right)\right].
\end{align*}

@@

\begin{figure}
\centering
%\subfigure{\includegraphics[scale=0.38, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig1_legend_01.pdf}}\\
%\vspace{-9mm}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 3mm 0mm, clip]{ch5_fig1_Cov_J.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig1_legend.pdf}}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 10mm 0mm, clip]{ch5_fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig2_Sigma_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 0mm 3mm 0mm, clip]{ch5_fig2_Cov_L.pdf}}
%\subfigure{\includegraphics[scale=0.38, trim = 29mm 48mm 124mm 10mm, clip]{ch5_fig2_legend.pdf}}
\caption{Mean posterior predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model (dashed red line) and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure}

Next, we analyze how the interaction effects between $m$ and $c$ affects the quality of the approximation. The lengthscale and marginal variance of the covariance function will no longer be fixed but instead their posterior marginal distributions estimated using dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual}, for both exact GP and HSGP models. Figure \ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the data for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated lengthscale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computations would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$  could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure \ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

@@

\begin{figure}
\centering
\begin{tabular}{ c c c }
\arrayrulecolor{darkgray}\hline
c = 1.05 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part1.pdf}
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part1.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part1.pdf} & 
\multirow{32}{1cm}{ \includegraphics[scale=0.35, trim = 28mm 30mm 100mm 30mm, clip]{ch5_fig3_legend.pdf}}\\ 
\arrayrulecolor{lightgray}\cline{1-2}
c = 1.1 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part2.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part2.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part2.pdf} &\\
\cline{1-2}
c = 1.2 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part3.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part3.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part3.pdf} &\\
\cline{1-2}
c = 1.5 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part4.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part4.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part4.pdf} & \\
\cline{1-2}
c = 2 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Post_part5.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{ch5_fig3_Sigma_part5.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{ch5_fig3_Cov_part5.pdf} & \\
\cline{1-2}
c = 2.5 &
\includegraphics[scale=0.215, trim = 0mm 4mm 0mm 14mm, clip]{ch5_fig3_Post_part6.pdf}
\includegraphics[scale=0.215, trim = 0mm 4mm 0mm 14mm, clip]{ch5_fig3_Sigma_part6.pdf}
\includegraphics[scale=0.215, trim = 0mm 4mm 5mm 14mm, clip]{ch5_fig3_Cov_part6.pdf} &\\
\arrayrulecolor{darkgray}\hline
\end{tabular}
\caption{Posterior mean predictive functions (left), covariance functions (center) and posterior standard deviations (right) of both the exact GP model and the HSGP model for different number of basis functions $m$ and for different values of the boundary factor $c$. The dashed red line of the exact GP can hardly be seen in the some of the plots because it is under the blue lines.}
  \label{fig3_Post_part1}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig5_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 28mm 119mm 0mm, clip]{ch5_fig5_legend.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure}

@@

In this example, we make a comparison of the lengthscale estimates
obtained from the exact GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.
For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both exact GP and HSGP models fitted to those datasets. The lengthscale estimates as obtained by exact GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As noted previously, an accurate estimate of the lengthscale can be a good indicator of a close approximation of the HSGP model to the exact GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the exact GP models, as a function of the lengthscale and number of basis functions.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Mean posterior predictions of both exact GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). The dashed red line of the exact GP can hardly be seen in the plots because it is almost perfectly overlayed by the blue lines.}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ccc}
\multicolumn{3}{c}{ \includegraphics[scale=0.42, trim = 25mm 117mm 90mm 45mm, clip]{ch5_fig8_legend.pdf}}\\
\includegraphics[scale=0.31, trim = 0mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part1.pdf} & \hspace{-4mm}
\includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part2.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part3.pdf} \\ 
\includegraphics[scale=0.31, trim = 0mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part4.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part5.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 0mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part6.pdf}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional lengthscales ($\ell$), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding lengthscale estimates $(\hat{\ell})$ from the exact GP and HSGP models. 95\% confident intervals of the lengthscale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

@@

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
where $I$ represents the identity matrix and $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function values to the noisy data. The previous formulation corresponds to the latent form of a GP model. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with a Mat{\'e}rn($\nu$=3/2) covariance function $k$. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$.

@@

In the HSGP model, the latent function values $f(x)$ are approximated as in equation (\ref{eq_approxf}), with the Mat{\'e}rn($\nu$=3/2) spectral density $S$ as in equation (\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively. 
In order to do model comparison, in addition to the exact GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the exact GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$; see equation (\ref{eq_boundary})), and the spline model with 80 knots. The true data-generative function and the noisy observations are also plotted. The sample observations are plotted as circles and the out-of-sample or test data, which have not been taking part on training the models, are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation. The posteriors of the three models, exact GP, HSGP and spline, are pretty similar within the interpolation input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{ch5_fig10_Posteriors_exI.pdf}
\caption{Posterior predictive means of the proposed HSGP model, the exact GP model, and the spline model. 95\% credible intervals are plotted as dashed lines.}
  \label{ch5_fig10_Posteriors_exI}
\end{figure}

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{ch5_fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{ch5_fig10_Posteriors_exI} and \ref{ch5_fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the exact GP model for interpolating and extrapolating data. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.
